Lossy compression of DNA sequencing quality data

### 2015-04-08 manuscript/q-compression.tex
Adjusted citations to accommodate the change from number to Harvard (author year) format. This was mostly changing \cite to \citep where appropriate. A few other small changes as well.

Quality values comprise a standard component of \textsc{fastq}
files~\cite{Cock:2010ve}, a very common format for sequence read data.
At the level of sequence read the probability of error for each
base-call is typically represented by \textsc{phred} quality value,
which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Quality values can be used throughout bioinformatics pipelines.  Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, the
elimination of reads that on the whole do not meet arbitrary quality
standards, which reduces the total number of reads; and
\textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in low
coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr},
although some recent work has addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.
Among the several challenges for compression of read data is dealing
with different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Sequence reads generated by instruments such
as an Illumina HiSeq, the focus of this research, are characterized by
having relatively few insertion and deletion errors, but substitution
(miscall) errors are much more frequent and have context-specific
patterns. These errors are non-uniformly distributed over the read
length (e.g., error rates increase up to $\sim$16$\times$ at the
3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
at the 3$^{\prime}$ end~\cite{Minoche:2011km}).

@@

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two, i.e., two
categories ``good'' and ``bad'' quality. We implement 2-bin encoding
by setting a quality value threshold empirically determined by the
distribution of quality values across reads. Base-calls are marked
``bad'' if their quality value falls below the first quartile minus
1.5 $\times$ the interquartile range (IQR), which is the difference
between the first and third quartile. 1.5 $\times$ IQR is the value
used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
of this approach is that it is completely data-dependent, and no
assumptions regarding the distribution of the quality values need to
be made.

With 2-bin encoding binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard 8
bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.

@@

$k$-means clustering is a vector quantization method, partitioning a
set of samples into $k$ sets that minimize within-cluster sum of
squares~\cite{macqueen1967some}. Using a random subset of read quality
values, a compression method can use the computed cluster centers as
read quality profiles. As the problem is NP-hard, we use a heuristic
iterative refinement approach quickly converging to a locally optimal
minimum provided by the Python package scikit-learn
0.15~\cite{pedregosa2011scikit}.

@@

We used several Illumina sequence read data sets in this research as
follows.

\textit{Rhodobacter sphaeroides} 2.4.1 data from the \textsc{gage}
(Genome Assembly Gold-Standard Evaluations)~\cite{Salzberg:2012rc},
which are generated from a fragment library (insert size of 180 nt;
2,050,868 paired-end reads) and short-jump library (insert size of
3,500 nt; 2,050,868 reads). The corresponding reference sequence was
obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1,
NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).

\textit{Stapylococcus aureus} USA300 data from the
\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
The corresponding reference sequence was obtained from the NCBI RefSeq
database (NC\_010063.1, NC\_010079.1, NC\_012417.1).

\textit{Homo sapiens} chromosome 14 data from the
\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
and short-jump library (insert sizes ranging from 2283-2803 nt;
22,669,408 reads).  The corresponding reference sequence was obtained
from the NCBI RefSeq database (NC\_000014.8).

@@

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\cite{sickle}, quantify read characteristics
with PrinSeq~\cite{Schmieder:2011gd}, and make comparison to
uncompressed data.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} that counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. For the genome assembly we used
software that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} \textsc{lap} metrics. The
5th-order polynomial regression assembles approximately 2.5 kb more
sequence and uses roughly $1.2\%$ more of the total read set than the
4th-order (Supplemental Table 1), which may explain the discrepency in
\textsc{lap} values because the \textsc{lap} value is influenced by
total coverage~\cite{Ghodsi:2013hb}.

@@

Certain short read alignment tools use the quality sequence
information when finding potential alignments. Bowtie2 (version 2.2.3)
was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie2 uses quality values written in the \textsc{fastq} files
when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
generated \textsc{sam} file for each compressing approach were
compared with the uncompressed \textsc{sam} file. The total, shared
and unique proportional numbers of mapped reads are calculated with
respect to the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was tweaked with the maximum and
minimum mismatch penalty equivalent to maximum and minimum quality
scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\cite{Kozanitis:2011kl} and
\textsc{QualComp}~\cite{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. Koren et al.~\cite{Koren:2013ye} have
shown that the assembly complexity of bacterial genomes can be greatly
simplified, producing near complete genome assemblies, by utilizing a
single run of these long reads. If long read sequencing technologies
such as PacBio become more widely adopted, it would be of huge benefit
to examine the potential of lossy compression algorithms on not only
the quality values, but the biological sequencing data themselves.

>>>>

Quality values comprise a standard component of \textsc{fastq}
files~\citep{Cock:2010ve}, a very common format for sequence read data.
At the level of sequence read the probability of error for each
base-call is typically represented by \textsc{phred} quality value,
which is defined as $Q =
-10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Quality values can be used throughout bioinformatics pipelines.  Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, the
elimination of reads that on the whole do not meet arbitrary quality
standards, which reduces the total number of reads; and
\textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie2~\citep{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in low
coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.  Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Sequence reads generated by instruments such
as an Illumina HiSeq, the focus of this research, are characterized by
having relatively few insertion and deletion errors, but substitution
(miscall) errors are much more frequent and have context-specific
patterns. These errors are non-uniformly distributed over the read
length (e.g., error rates increase up to $\sim$16$\times$ at the
3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
at the 3$^{\prime}$ end~\citep{Minoche:2011km}).

@@

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two, i.e., two
categories ``good'' and ``bad'' quality. We implement 2-bin encoding
by setting a quality value threshold empirically determined by the
distribution of quality values across reads. Base-calls are marked
``bad'' if their quality value falls below the first quartile minus
1.5 $\times$ the interquartile range (IQR), which is the difference
between the first and third quartile. 1.5 $\times$ IQR is the value
used by Tukey's box plot~\citep{mcgill1978variations}. The main
benefit of this approach is that it is completely data-dependent, and
no assumptions regarding the distribution of the quality values need
to be made.

With 2-bin encoding binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard
8 bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78}.

@@

$k$-means clustering is a vector quantization method, partitioning a
set of samples into $k$ sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. Using a random subset of read quality
values, a compression method can use the computed cluster centers as
read quality profiles. As the problem is NP-hard, we use a heuristic
iterative refinement approach quickly converging to a locally optimal
minimum provided by the Python package scikit-learn
0.15~\citep{pedregosa2011scikit}.

@@

We used several Illumina sequence read data sets in this research,
which are taken from data from the \textsc{gage} (Genome Assembly
Gold-Standard Evaluations)~\citep{Salzberg:2012rc} except as
noted. These data sets are as follows.

\textit{Rhodobacter sphaeroides} 2.4.1, which are generated from a
fragment library (insert size of 180 nt; 2,050,868 paired-end reads)
and short-jump library (insert size of 3,500 nt; 2,050,868 reads). The
corresponding reference sequence was obtained from the NCBI RefSeq
database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1,
NC\_007494.1, NC\_009007.1, NC\_009008.1).

\textit{Stapylococcus aureus} USA300, which are generated from a
fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
The corresponding reference sequence was obtained from the NCBI RefSeq
database (NC\_010063.1, NC\_010079.1, NC\_012417.1).

\textit{Homo sapiens} chromosome 14 data, which are generated from a
fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
and short-jump library (insert sizes ranging from 2283-2803 nt;
22,669,408 reads).  The corresponding reference sequence was obtained
from the NCBI RefSeq database (NC\_000014.8).

@@

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\citep{sickle}, quantify read characteristics
with PrinSeq~\citep{Schmieder:2011gd}, and make comparison to
uncompressed data.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} that counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. For the genome assembly we used
software that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} \textsc{lap} metrics. The
5th-order polynomial regression assembles approximately 2.5 kb more
sequence and uses roughly $1.2\%$ more of the total read set than the
4th-order (Supplemental Table 1), which may explain the discrepency in
\textsc{lap} values because the \textsc{lap} value is influenced by
total coverage~\citep{Ghodsi:2013hb}.

@@

Certain short read alignment tools use the quality sequence
information when finding potential alignments. Bowtie2 (version 2.2.3)
was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie2 uses quality values written in the \textsc{fastq} files
when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
generated \textsc{sam} file for each compressing approach were
compared with the uncompressed \textsc{sam} file. The total, shared
and unique proportional numbers of mapped reads are calculated with
respect to the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{QualComp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.
